---
title: "Practical Machine Learning Course Project Write-Up"
author: "Chelsea Heaven"
date: "Saturday, October 17, 2015"
output: html_document
---

#Introduction

Hello Git! This is the first test commit of my project for the practical machine learning  class.


```{r retreivedata, echo=FALSE}
#Obtaining data for class project
urltrain <- c("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
urltest <- c("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")

download.file(urltrain,destfile="ProjectTrainingSet.csv")
download.file(urltest,destfile="ProjectTestSet.csv")

tst <- read.csv("ProjectTestSet.csv")
train  <- read.csv("ProjectTrainingSet.csv")

```

```{r graph, echo=FALSE}

#Look at this graaaaph
plot(mtcars)

```



#Model Details

##Building the Model

##Cross Validation of the Model

##Expected Sample Error

##Justification Of Model Choices

#Prediction of 20 Test Cases

#Conclusion

##Does the submission build a machine learning algorithm to predict activity quality from activity monitors?

##Do the authors describe what they expect the out of sample error to be and estimate the error appropriately with cross-validation?

